---
title: "Math 460: Exam 1"
# subtitle: "Semester Fall 2026"
# date: "December 15, 2026"
# author: "Department of Mathematics"
format:
  pdf:
    documentclass: article
    geometry:
      - top=.5in
      - bottom=.5in
      - left=.5in
      - right=.5in
    header-includes:
      - \pagestyle{empty}
      - \pagenumbering{gobble}
  html:
    toc: false
    theme: cosmo
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

**Student Name:** \underline{\hspace{8cm}} 

## Problem

```{r, echo = FALSE}
library(tidyverse)
ip = function(n0, n1){
  p0 = n0/(n0+n1)
  return(1-p0^2-(1-p0)^2)}

ig = function(a0, a1, b0, b1){
  k = (b0+b1)/(a0+a1)
  return(ip(a0, a1) - k*ip(b0,b1)-(1-k)*ip(a0-b0, a1-b1))}

 ig2 = function(a0, a1, b0, b1)
 {
  n = a0+a1+b0+b1
  i_left  = ip(a0, a1) 
  i_right = ip(b0, b1)
  i_parent = ip(a0+b0, a1+b1)
  i_children = ((a0+a1)/n)*i_left + ((b0+b1)/n)*i_right
  i_gain =   i_parent-i_children
  return(i_gain)
 } 

igg = function(df, v, s)
# df = Data
# v = Index of the Variable to split
# s = Value the variable split at 
  
{
  #set the variable (x1) and the value it splits at(x1==1)
  s1 = table(df[df[,v]==s,]$target)
  tt = table(df$target)
  #calculate the impurity gain of the split
  ig1 = ig(tt[1],tt[2],s1[1],s1[2])
  
  return(ig1)
  
}
ID =       c(1,2,3,4,5,6,7,8,9, 10,11,12,13,14,15,16)
Class     =c(1,1,1,1,1,1,1,3,2,2,3,2,3,1,2,1)
Sex     =  c(0,0,0,1,0,0,0,1,1,1,1,1,0,0,0,1)
target =   c(1,1,1,1,1,1,1,1,0,0,0,0,0,0,0,0)
d = as_tibble(cbind(ID, Class, Sex, target))

```

-   Given the **training** data with the binary target variable `Survived` and two input categorical variables `Class` and `Sex`. (Notice: ID is not a variable of interest). 

<center>

```{r, echo = FALSE}
names(d) = c("ID", "Class", "Sex", "Survived")
knitr::kable(d)
```

</center>

Using Gini Index as the measure for impurity to:

1.  Grow the **maximum tree** with three leaves (a stopping rule!) on the data. Draw the (diagram of) tree. 

  - Notice: If two splits achieve the same maximum impurity gain, you may randomly select either split as the best split.

2.  Find the misclassification rate on training data of the maximal tree
3.  Draw all the possible **subtrees**
4.  Validate the maximal tree and all the subtrees on the following data to select the **optimal tree**. 

  - Note: If the chance of Survived is 1/2, predict `Survived` (`Survived`=1)

<center>

```{r, echo = FALSE}
knitr::kable(as_tibble(data.frame(Class=c(1,1,3,2), Sex =c(0,1,1,0), Survived =c(1,1,0,0))))
```

</center>